---
title: "Module 1: Introduction to BRFSS and Binge Drinking Epidemiology"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: "flatly"
runtime: shiny_prerendered
description: "Learn what BRFSS is, how binge drinking is defined, and what this analysis aims to accomplish."
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
tutorial_options(exercise.timelimit = 60)
```

## Welcome to the Course

This course teaches you to conduct a **national-scale epidemiological surveillance analysis** of binge drinking trends in the United States using the Behavioral Risk Factor Surveillance System (BRFSS). You will work through the same methodology used in peer-reviewed research — from raw survey data through regression modeling and publication-quality visualization.

**What you will learn:**

- Load and harmonize 14 years of BRFSS survey data (2011–2024)
- Handle the complexities of nationally representative complex survey designs
- Compute age-standardized prevalence estimates with confidence intervals
- Stratify analyses by sex, race/ethnicity, education, and income
- Model trends with Poisson regression and test for effect modification
- Build publication-ready figures in ggplot2

**Prerequisites:** Familiarity with R, dplyr, and ggplot2 basics.

---

## What is BRFSS?

The **Behavioral Risk Factor Surveillance System (BRFSS)** is the world's largest continuously conducted health survey system.

**Key facts:**

| Feature | Detail |
|---------|--------|
| Sponsor | U.S. Centers for Disease Control and Prevention (CDC) |
| Frequency | Annual (every year since 1984) |
| Coverage | All 50 U.S. states + Washington D.C. + territories |
| Respondents | ~450,000 adults (18+) per year |
| Mode | Telephone survey (landline + cellular) |
| Data format | SAS transport (XPT) files |

**Why BRFSS?**

BRFSS is the primary data source for state-level behavioral health surveillance in the U.S. Because it samples every state annually, it allows us to:

1. Track **national trends** over time
2. Compare **state-to-state variation**
3. Estimate prevalence in **demographic subgroups** with adequate sample sizes
4. Make **population-representative** inferences through survey weighting

**Limitations to keep in mind:**

- Self-reported data (recall bias, social desirability bias)
- Excludes institutionalized populations and those without phones
- Variable names and coding **change across years** — careful harmonization required
- Response rates have declined over time (~50% in recent years)

---

## The CDC Definition of Binge Drinking

The CDC defines **binge drinking** as a pattern of drinking that brings blood alcohol concentration (BAC) to 0.08 g/dL or higher. In practice, this corresponds to consuming:

- **5 or more drinks** within about 2 hours for **men**
- **4 or more drinks** within about 2 hours for **women**

BRFSS captures this through two approaches:

**Approach 1 — CDC calculated flag (`_RFBING5`/`_RFBING6`):**
A derived binary variable the CDC computes from raw responses. Values: `1 = No`, `2 = Yes` (note the reverse coding!).

**Approach 2 — Raw episodes (`DRNK3GE5`):**
"During the past 30 days, how many times did you have 5 or more drinks on an occasion?" (The threshold is 5 for everyone — sex-specific adjustment happens in the CDC flag derivation.)

In this analysis, **we prefer the CDC flag** (`_RFBING5`) and fall back to episode counts only when the flag is unavailable.

```{r binge-def-quiz}
quiz(
  question(
    "For a woman, how many drinks within ~2 hours qualifies as binge drinking by CDC definition?",
    answer("3 or more"),
    answer("4 or more", correct = TRUE),
    answer("5 or more"),
    answer("6 or more"),
    allow_retry = TRUE
  ),
  question(
    "In the BRFSS _RFBING5 variable, what does the value '2' mean?",
    answer("No binge drinking"),
    answer("Yes, binge drinking", correct = TRUE),
    answer("Missing / unknown"),
    answer("Refused to answer"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  )
)
```

---

## BRFSS Survey Design Basics

BRFSS uses a **stratified cluster sampling** design. Understanding this is critical — **you cannot use simple R functions like `mean()` or `lm()` on BRFSS data without accounting for the survey design.** Doing so produces incorrect standard errors and biased estimates.

**Three components define the survey design:**

### 1. Survey Weights (`_LLCPWT`)
Each respondent is assigned a weight that represents how many people in the population they "stand for." Weights account for:

- Unequal probability of selection (e.g., households with only cell phones)
- Nonresponse adjustment
- Post-stratification to census demographics (age, sex, race/ethnicity)

### 2. Strata (`_STSTR`)
BRFSS divides each state into geographic strata (usually urban/rural sub-regions). Sampling happens within strata to ensure geographic coverage.

### 3. Primary Sampling Units — PSUs (`_PSU`)
Within each stratum, telephone numbers are grouped into clusters (PSUs). PSUs are the primary unit of selection.

**The `survey` package in R** handles all of this correctly. You specify the design once, then use survey-aware versions of common functions.

```{r survey-design-quiz}
quiz(
  question(
    "Which R function should you use to compute a mean from BRFSS data?",
    answer("mean()", message = "This ignores survey weights and complex design — it gives wrong answers."),
    answer("svymean()", correct = TRUE, message = "Correct! svymean() uses the survey design to produce properly weighted and variance-corrected estimates."),
    answer("weighted.mean()", message = "weighted.mean() handles weights but not the complex variance from clustering and stratification."),
    answer("summarise() with wt = weight", message = "This handles weights but not clustering — standard errors will be wrong."),
    allow_retry = TRUE
  )
)
```

---

## Overview of the Analysis

This course walks through a 14-year trend analysis of binge drinking in the U.S.:

**Study population:** U.S. adults aged 18+ in all 50 states and D.C.

**Time period:** 2011–2024

**Primary outcome:** Any binge drinking (CDC-defined binary flag)

**Secondary outcomes:**
- Binge drinking among current drinkers only
- High-intensity binge drinking (≥8 drinks for women, ≥10 for men)
- Binge episode frequency categories

**Key analyses:**
1. Overall age-adjusted prevalence by year (trend)
2. Prevalence stratified by sex, race/ethnicity, education, income
3. State-level estimates and rankings
4. Adjusted Prevalence Ratio (PR) regression models
5. Race × Time and Sex × Time interaction tests

**Data flow:**

```
Raw XPT files (2011–2024)
        ↓
  Load & audit variable availability
        ↓
  Harmonize variable names across years
        ↓
  Recode outcomes & covariates
        ↓
  Build complex survey design objects
        ↓
  Age-standardize (Year-2000 standard)
        ↓
  Compute prevalence estimates + CIs
        ↓
  Regression models
        ↓
  Export tables + figures
```

---

## R Package Overview

The analysis relies on these core packages:

```{r packages, exercise=TRUE}
# Run this to see which packages we'll use
pkgs <- c(
  "tidyverse",  # data manipulation + ggplot2
  "haven",      # read SAS XPT files
  "janitor",    # clean variable names
  "survey",     # complex survey methods
  "srvyr",      # tidyverse-style survey interface
  "scales",     # axis formatting in ggplot2
  "purrr"       # functional programming (map/map_dfr)
)

# Check which are installed
installed <- pkgs %in% rownames(installed.packages())
data.frame(
  package   = pkgs,
  installed = installed,
  role      = c(
    "Data manipulation + visualization",
    "Read SAS XPT files",
    "Clean column names",
    "Complex survey estimation",
    "Tidyverse-style survey API",
    "ggplot2 axis labels",
    "Functional iteration"
  )
)
```

---

## Key Option: Lonely PSUs

Before running any survey analysis, set this option:

```{r lonely-psu, exercise=TRUE}
# This handles strata where only one PSU was sampled
# (common at the edges of the BRFSS sample)
options(survey.lonely.psu = "adjust")

# What it does:
cat("survey.lonely.psu = 'adjust'\n")
cat("When a stratum has only one PSU, its contribution to variance\n")
cat("is computed using the overall mean — rather than causing an error.\n")
cat("This is the recommended approach for BRFSS.\n")
```

```{r lonely-psu-quiz}
quiz(
  question(
    "Why do we set options(survey.lonely.psu = 'adjust') at the start of a BRFSS analysis?",
    answer("To speed up computation"),
    answer("To prevent errors when a stratum has only one sampled PSU", correct = TRUE),
    answer("To enable parallel processing"),
    answer("To suppress missing value warnings"),
    allow_retry = TRUE
  )
)
```

---

## Module Summary

You have learned:

- **BRFSS** is a nationally representative annual telephone survey of U.S. adults, covering all 50 states + D.C.
- **Binge drinking** is defined as ≥5 drinks (men) or ≥4 drinks (women) in ~2 hours
- BRFSS uses a **stratified cluster design** requiring the `survey` package for valid inference
- The three design components are: **weights** (`_LLCPWT`), **strata** (`_STSTR`), and **PSUs** (`_PSU`)
- Variable names and coding **change across survey years** — harmonization is essential

**Next:** Module 2 covers how to load BRFSS XPT files and audit variable availability across years.
