---
title: "Module 3: Data Harmonization Across Years"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: "flatly"
runtime: shiny_prerendered
description: "Recode raw BRFSS variables into consistent, analysis-ready categories across 2011–2024."
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(janitor)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
tutorial_options(exercise.timelimit = 60)

source("synthetic_brfss.R", local = TRUE)
df_raw <- make_synthetic_brfss(n_per_year = 1000, years = 2011:2024)

# Helper for exercises
as_int <- function(x) suppressWarnings(as.integer(as.character(x)))
```

## Why Harmonization is Necessary

Even after resolving variable *names*, the variable *values* often change across BRFSS years:

- **Income:** early years used `INCOME2` (8 categories), later years added `INCOME3` (11 categories)
- **Race/ethnicity:** multiple source variables with overlapping and changing codes
- **Sex:** response options expanded; older "male/female" coding differs from newer versions
- **Alcohol variables:** `ALCDAY5` (days/week or days/month) was renamed `ALCDAY4` in 2020

The goal of harmonization is to create a **single consistent set of analysis variables** that works across all years, using a hierarchical fallback strategy where preferred variables are not available.

---

## Outcome Variable: Any Binge Drinking

The primary outcome requires two approaches:

**Approach 1 (preferred):** The CDC pre-computed flag `_RFBING5` / `_RFBING6`
- Values: `1 = No binge drinking`, `2 = Yes binge drinking`
- Note the **reverse coding** — 1 = No, 2 = Yes

**Approach 2 (fallback):** Derived from binge episode count `DRNK3GE5`
- Values: `0–76 = number of episodes`, `88 = zero times`, `77/99 = DK/Refused`

```{r binge-outcome-def}
alcday5_to_days30 <- function(x) {
  xi <- suppressWarnings(as.integer(as.character(x)))
  ifelse(is.na(xi), NA_real_,
    ifelse(xi == 888L, 0,
      ifelse(xi >= 101L & xi <= 107L, (xi - 100) * 4.3,   # days/week × ~4.3
        ifelse(xi >= 201L & xi <= 231L, (xi - 200),        # days/month
          NA_real_))))
}

recode_binge_outcome <- function(d) {
  d |>
    dplyr::mutate(
      # Step 1: clean binge episode count
      binge_episodes_clean = dplyr::case_when(
        !is.na(drnk3ge5) & drnk3ge5 %in% 1:76  ~ as.integer(drnk3ge5),  # actual count
        !is.na(drnk3ge5) & drnk3ge5 == 88       ~ 0L,                   # "none"
        !is.na(drnk3ge5) & drnk3ge5 %in% c(77, 99) ~ NA_integer_,       # DK/Refused
        TRUE ~ NA_integer_
      ),

      # Step 2: fill zero from alcday5 if episode still missing
      binge_episodes = dplyr::case_when(
        !is.na(binge_episodes_clean)        ~ binge_episodes_clean,
        is.na(binge_episodes_clean) & alcday5 == 888 ~ 0L,  # non-drinker
        TRUE ~ NA_integer_
      ),

      # Step 3: primary outcome from CDC flag (reverse-coded: 1=No, 2=Yes)
      any_binge_cdc = dplyr::case_when(
        !is.na(rfbing5) & as.integer(rfbing5) == 2 ~ 1L,  # Yes
        !is.na(rfbing5) & as.integer(rfbing5) == 1 ~ 0L,  # No
        TRUE ~ NA_integer_
      ),

      # Step 4: fallback to episodes if CDC flag missing
      any_binge_cdc = dplyr::if_else(
        is.na(rfbing5),
        dplyr::case_when(
          is.na(alcday5)          ~ NA_integer_,
          binge_episodes > 0      ~ 1L,
          binge_episodes == 0     ~ 0L,
          TRUE                    ~ NA_integer_
        ),
        any_binge_cdc
      )
    )
}
```

```{r binge-outcome-quiz}
quiz(
  question(
    "In the BRFSS _RFBING5 variable, what does value '1' mean?",
    answer("Yes, respondent binge drinks"),
    answer("No binge drinking", correct = TRUE),
    answer("Refused to answer"),
    answer("Don't know"),
    allow_retry = TRUE,
    message = "BRFSS reverse-codes this: 1 = No, 2 = Yes. Easy to miss!"
  )
)
```

---

## Exercise: Recode Binge Outcome

```{r binge-recode-exercise, exercise=TRUE}
# Apply the binge outcome recode to synthetic data
df_sample <- df_raw |>
  dplyr::rename(
    drnk3ge5 = drnk3ge5,   # already named correctly in synthetic data
    rfbing5  = rfbing5
  ) |>
  dplyr::slice_head(n = 20)

result <- recode_binge_outcome(df_sample)

# Check distribution of any_binge_cdc
result |>
  dplyr::count(rfbing5, any_binge_cdc) |>
  print()
```

```{r binge-freq-recode, exercise=TRUE}
# Recode binge episodes into frequency categories
recode_freq5 <- function(episodes) {
  dplyr::case_when(
    is.na(episodes)          ~ NA_character_,
    episodes == 0            ~ "none (0)",
    episodes == 1            ~ "monthly (1)",
    episodes %in% 2:3        ~ "a few times monthly (2-3)",
    episodes %in% 4:7        ~ "weekly (4-7)",
    episodes >= 8            ~ ">=2x/week (>=8)"
  )
}

df_raw |>
  dplyr::mutate(freq5 = recode_freq5(drnk3ge5)) |>
  dplyr::count(freq5, sort = TRUE)
```

---

## Age Group Harmonization

BRFSS uses `_AGEG5YR` (codes 1–13 for 5-year age bands starting at 18–24). We map these to two grouping schemes:

**10-bin primary:** 18–24, 25–29, ..., 60–64, 65+

**7-band (for age standardization):** 18–29, 30–39, 40–49, 50–59, 60–69, 70–79, 80+

```{r age-recode-def}
recode_age <- function(age5_col) {
  age5 <- suppressWarnings(as.integer(as.character(age5_col)))
  list(
    age_group = dplyr::case_when(
      age5 == 1            ~ "18-24",
      age5 == 2            ~ "25-29",
      age5 == 3            ~ "30-34",
      age5 == 4            ~ "35-39",
      age5 == 5            ~ "40-44",
      age5 == 6            ~ "45-49",
      age5 == 7            ~ "50-54",
      age5 == 8            ~ "55-59",
      age5 == 9            ~ "60-64",
      age5 %in% 10:13      ~ "65+",
      TRUE                 ~ NA_character_
    ),
    age7 = dplyr::case_when(
      age5 %in% 1:2        ~ "18-29",
      age5 %in% 3:4        ~ "30-39",
      age5 %in% 5:6        ~ "40-49",
      age5 %in% 7:8        ~ "50-59",
      age5 %in% 9:10       ~ "60-69",
      age5 %in% 11:12      ~ "70-79",
      age5 == 13           ~ "80+",
      TRUE                 ~ NA_character_
    )
  )
}
```

```{r age-recode-exercise, exercise=TRUE}
# Recode age groups and check distribution
age_recoded <- recode_age(df_raw$age5)

df_raw |>
  dplyr::mutate(
    age_group = age_recoded$age_group,
    age7      = age_recoded$age7
  ) |>
  dplyr::count(age_group) |>
  print()
```

```{r age7-exercise, exercise=TRUE}
age_recoded <- recode_age(df_raw$age5)
df_raw |>
  dplyr::mutate(age7 = age_recoded$age7) |>
  dplyr::count(age7)
```

---

## Race/Ethnicity Harmonization

This is the **most complex** harmonization step. BRFSS has multiple race variables that changed over time:

| Variable | Years available | Description |
|----------|----------------|-------------|
| `_IMPRACE` | Most years | CDC-imputed race (1–6) |
| `RACE1`/`_RACE1` | Older years | Single race, 7 categories |
| `_MRACE1` | 2015–2016 | Multiracial with detailed breakdown |
| `HISPANC`/`HISPANC2` | All years | Hispanic origin (separate question) |

**Critical principle:** Hispanic ethnicity is asked as a **separate question** from race. We always code Hispanic origin first (regardless of race responses) per CDC/OMB standards.

```{r race-recode-def}
recode_race <- function(d) {
  as_int_local <- function(x) suppressWarnings(as.integer(as.character(x)))

  imprace_i <- as_int_local(d$imprace)
  hisp_i    <- as_int_local(d$hispanc)
  race1_i   <- as_int_local(d$race1)

  race4 <- dplyr::case_when(
    # HISPANIC FIRST — always from HISPANC, not IMPRACE
    !is.na(hisp_i) & hisp_i == 1L              ~ "Hispanic",

    # NH categories from _IMPRACE (preferred)
    !is.na(imprace_i) & imprace_i == 1L        ~ "NH White",
    !is.na(imprace_i) & imprace_i == 2L        ~ "NH Black",
    !is.na(imprace_i) & imprace_i %in% 3:7     ~ "NH Other",

    # Fallback to RACE1 when _IMPRACE is missing
    is.na(imprace_i) & !is.na(race1_i) & race1_i == 1L ~ "NH White",
    is.na(imprace_i) & !is.na(race1_i) & race1_i == 2L ~ "NH Black",
    is.na(imprace_i) & !is.na(race1_i) & race1_i %in% 3:7 ~ "NH Other",

    TRUE ~ NA_character_
  )

  race4
}
```

```{r race-exercise, exercise=TRUE}
# Apply race harmonization to synthetic data
# Note: synthetic data uses 'imprace' column directly
df_race <- df_raw |>
  dplyr::mutate(
    imprace  = imprace,          # _IMPRACE codes 1-6
    hispanc  = NA_integer_,      # no hispanic variable in synthetic data
    race1    = NA_integer_       # no race1 fallback in synthetic data
  )

race4_result <- recode_race(df_race)
table(race4_result, useNA = "ifany")
```

---

## Education Harmonization

BRFSS has two education variables:
- `EDUCAG` (4 categories, CDC-derived): 1=<HS, 2=HS/GED, 3=Some college/AA, 4=College
- `EDUCA` (6 categories, raw): 1–3=grades 1–11, 4=HS grad, 5=Some college, 6=College grad

We prefer `EDUCAG` and fall back to collapsing `EDUCA`:

```{r educ-recode-def}
recode_educ4 <- function(educag_col, educa_col = NULL) {
  as_int_local <- function(x) suppressWarnings(as.integer(as.character(x)))

  educag_i <- as_int_local(educag_col)
  educa_i  <- if (!is.null(educa_col)) as_int_local(educa_col) else rep(NA_integer_, length(educag_i))

  dplyr::case_when(
    # Prefer EDUCAG (CDC 4-level)
    !is.na(educag_i) ~ dplyr::recode(educag_i,
      `1` = "<HS", `2` = "HS/GED",
      `3` = "Some college/AA", `4` = "College",
      .default = NA_character_
    ),
    # Fallback: collapse EDUCA (6-level) into same 4 groups
    is.na(educag_i) & !is.na(educa_i) ~ dplyr::case_when(
      educa_i %in% 1:3 ~ "<HS",
      educa_i == 4     ~ "HS/GED",
      educa_i == 5     ~ "Some college/AA",
      educa_i == 6     ~ "College",
      TRUE             ~ NA_character_
    ),
    TRUE ~ NA_character_
  )
}
```

```{r educ-exercise, exercise=TRUE}
educ4_result <- recode_educ4(df_raw$educag)

table(educ4_result, useNA = "ifany")
```

---

## Income Harmonization

Income is the trickiest harmonization due to multiple source variables with different scales:

- `INCOME3` (2013+): 11 categories up to ≥$200k
- `INCOME2` (all years): 8 categories up to ≥$75k
- `_INCOMG` (some years): 6 categories

We create **three groupings** for different analytical purposes:

```{r income-recode-def}
recode_income3 <- function(income3_col, income2_col, incomg_col = NULL) {
  as_int_local <- function(x) suppressWarnings(as.integer(as.character(x)))

  income3_i <- as_int_local(income3_col)
  income2_i <- as_int_local(income2_col)
  incomg_i  <- if (!is.null(incomg_col)) as_int_local(incomg_col) else rep(NA_integer_, length(income3_i))

  dplyr::case_when(
    # DK/Refused from any source
    !is.na(income3_i) & income3_i %in% c(77, 99) ~ "Unknown",
    !is.na(income2_i) & income2_i %in% c(77, 99) ~ "Unknown",

    # Prefer INCOME3 (most granular)
    !is.na(income3_i) & income3_i %in% 1:4  ~ "<$25k",
    !is.na(income3_i) & income3_i %in% 5:6  ~ "$25-<50k",
    !is.na(income3_i) & income3_i %in% 7:11 ~ ">=50k",

    # Fallback to INCOME2
    is.na(income3_i) & !is.na(income2_i) & income2_i %in% 1:4 ~ "<$25k",
    is.na(income3_i) & !is.na(income2_i) & income2_i %in% 5:6 ~ "$25-<50k",
    is.na(income3_i) & !is.na(income2_i) & income2_i %in% 7:8 ~ ">=50k",

    # Fallback to _INCOMG
    is.na(income3_i) & is.na(income2_i) & !is.na(incomg_i) & incomg_i %in% 1:2 ~ "<$25k",
    is.na(income3_i) & is.na(income2_i) & !is.na(incomg_i) & incomg_i %in% 3:4 ~ "$25-<50k",
    is.na(income3_i) & is.na(income2_i) & !is.na(incomg_i) & incomg_i %in% 5:6 ~ ">=50k",

    TRUE ~ "Unknown"
  )
}
```

```{r income-exercise, exercise=TRUE}
income3_result <- recode_income3(
  income3_col = NA_integer_,   # not in synthetic data
  income2_col = df_raw$income2
)

table(income3_result, useNA = "ifany")
```

---

## Putting It Together: `recode_analysis()`

In the actual analysis, all recoding is applied in a single `recode_analysis()` function. This ensures consistency and makes the pipeline reproducible.

```{r recode-pipeline, exercise=TRUE}
# Demonstrate the recoding pipeline on a small slice
source("synthetic_brfss.R")
df_sample <- make_synthetic_brfss(n_per_year = 200, years = 2019:2021)

# Apply a simplified version of the full recode
df_recoded <- df_sample |>
  dplyr::mutate(
    # Binge outcome
    any_binge_cdc = dplyr::case_when(
      rfbing5 == 2 ~ 1L,
      rfbing5 == 1 ~ 0L,
      TRUE ~ NA_integer_
    ),

    # Age groups
    age_group = dplyr::case_when(
      age5 == 1   ~ "18-24", age5 == 2 ~ "25-29", age5 == 3 ~ "30-34",
      age5 == 4   ~ "35-39", age5 == 5 ~ "40-44", age5 == 6 ~ "45-49",
      age5 == 7   ~ "50-54", age5 == 8 ~ "55-59", age5 == 9 ~ "60-64",
      age5 %in% 10:13 ~ "65+", TRUE ~ NA_character_
    ) |> factor(levels = c("18-24","25-29","30-34","35-39","40-44",
                            "45-49","50-54","55-59","60-64","65+"), ordered = TRUE),

    # Education
    educ4 = dplyr::recode(as.integer(educag),
      `1`="<HS", `2`="HS/GED", `3`="Some college/AA", `4`="College",
      .default = NA_character_
    ) |> factor(levels = c("<HS","HS/GED","Some college/AA","College"), ordered = TRUE),

    # Sex
    sex_f = factor(as.integer(sex), levels = c(2, 1), labels = c("Female", "Male"))
  )

cat("=== Variable summary after recoding ===\n\n")
cat("Years:", paste(sort(unique(df_recoded$year)), collapse = ", "), "\n")
cat("N rows:", nrow(df_recoded), "\n\n")

cat("Binge outcome:\n")
print(table(df_recoded$any_binge_cdc, useNA = "ifany"))

cat("\nAge groups:\n")
print(table(df_recoded$age_group))

cat("\nEducation:\n")
print(table(df_recoded$educ4))
```

---

## Module Summary

You have learned:

- **Outcome coding:** The CDC binge flag is reverse-coded (1=No, 2=Yes); always use it as primary with episode-count fallback
- **Age harmonization:** `_AGEG5YR` codes 1–13 map to 10-bin or 7-band age groups
- **Race/ethnicity:** Hispanic identity (from separate HISPANC question) takes precedence over `_IMPRACE` race codes
- **Education:** Prefer `EDUCAG` (4-level), fall back to collapsing `EDUCA` (6-level) into the same groups
- **Income:** Multiple source variables across years require a 3-level fallback hierarchy
- All recoding is centralized in `recode_analysis()` for consistency and reproducibility

**Next:** Module 4 covers building proper complex survey design objects in R.
