---
title: "Module 4: Complex Survey Design in R"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: "flatly"
runtime: shiny_prerendered
description: "Build correct survey design objects using the survey package and understand why it matters."
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(survey)
library(srvyr)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
tutorial_options(exercise.timelimit = 90)
options(survey.lonely.psu = "adjust")

source("synthetic_brfss.R", local = TRUE)

# Pre-build recoded data for exercises
df_raw <- make_synthetic_brfss(n_per_year = 2000, years = 2011:2024)

df_all <- df_raw |>
  dplyr::mutate(
    any_binge_cdc = dplyr::case_when(
      rfbing5 == 2 ~ 1L, rfbing5 == 1 ~ 0L, TRUE ~ NA_integer_
    ),
    sex_f = factor(sex, levels = c(2, 1), labels = c("Female", "Male")),
    age_group = dplyr::case_when(
      age5 == 1 ~ "18-24", age5 == 2 ~ "25-29", age5 == 3 ~ "30-34",
      age5 == 4 ~ "35-39", age5 == 5 ~ "40-44", age5 == 6 ~ "45-49",
      age5 == 7 ~ "50-54", age5 == 8 ~ "55-59", age5 == 9 ~ "60-64",
      age5 %in% 10:13 ~ "65+", TRUE ~ NA_character_
    ) |> factor(levels = c("18-24","25-29","30-34","35-39","40-44",
                            "45-49","50-54","55-59","60-64","65+"), ordered = TRUE),
    age7 = dplyr::case_when(
      age5 %in% 1:2 ~ "18-29", age5 %in% 3:4 ~ "30-39",
      age5 %in% 5:6 ~ "40-49", age5 %in% 7:8 ~ "50-59",
      age5 %in% 9:10 ~ "60-69", age5 %in% 11:12 ~ "70-79",
      age5 == 13 ~ "80+", TRUE ~ NA_character_
    ) |> factor(levels = c("18-29","30-39","40-49","50-59","60-69","70-79","80+")),
    educ4 = dplyr::recode(as.integer(educag),
      `1`="<HS", `2`="HS/GED", `3`="Some college/AA", `4`="College",
      .default = NA_character_
    ) |> factor(levels = c("<HS","HS/GED","Some college/AA","College"), ordered = TRUE),
    strata_id = as.integer(interaction(year, strata, drop = TRUE)),
    psu_id    = as.integer(interaction(year, psu,    drop = TRUE)),
    year_lin  = as.numeric(year) - 2011,
    is_current = dplyr::case_when(
      alcday5 %in% c(777, 999) ~ NA,
      alcday5 == 888 ~ FALSE,
      TRUE ~ TRUE
    )
  ) |>
  dplyr::filter(!is.na(sex_f), !is.na(age_group))

des_all <- survey::svydesign(
  id      = ~psu_id,
  strata  = ~strata_id,
  weights = ~wt,
  data    = df_all,
  nest    = TRUE
)
```

## Why Simple Statistics Are Wrong for BRFSS

The most common mistake when analyzing survey data is using unweighted statistics or ignoring the complex design.

**Let's see the effect directly:**

```{r wrong-vs-right, exercise=TRUE}
# Compare: wrong (ignores design) vs. correct (survey-weighted)
df_year <- df_all |> dplyr::filter(year == 2019)

# WRONG: simple unweighted mean
wrong_mean <- mean(df_year$any_binge_cdc, na.rm = TRUE)

# WRONG: weighted mean (correct point estimate but wrong variance)
wrong_weighted <- weighted.mean(df_year$any_binge_cdc, df_year$wt, na.rm = TRUE)

# CORRECT: survey design
des_yr <- survey::svydesign(
  id      = ~psu_id,
  strata  = ~strata_id,
  weights = ~wt,
  data    = df_year,
  nest    = TRUE
)
correct_est <- survey::svymean(~any_binge_cdc, des_yr, na.rm = TRUE)

cat("Unweighted mean:     ", round(wrong_mean, 4), "\n")
cat("Weighted mean:       ", round(wrong_weighted, 4), "\n")
cat("Survey estimate:     ", round(coef(correct_est), 4), "\n")
cat("Survey SE:           ", round(SE(correct_est), 4), "\n\n")
cat("Note: weighted.mean() gets the point estimate right but the SE is wrong.\n")
cat("Only svymean() uses the correct variance formula accounting for clustering.\n")
```

---

## Building a Survey Design Object

The core function is `survey::svydesign()`. You must specify:

```r
des <- survey::svydesign(
  id      = ~psu_id,      # PSU (cluster identifier)
  strata  = ~strata_id,   # Stratum identifier
  weights = ~wt,          # Survey weight
  data    = df,
  nest    = TRUE          # PSU IDs are nested within strata (not globally unique)
)
```

**The `nest = TRUE` argument is critical.** BRFSS PSU identifiers may repeat across strata (PSU "1" in stratum A is different from PSU "1" in stratum B). Setting `nest = TRUE` tells the `survey` package to treat PSU IDs as nested within strata.

---

## Pooling Years: Making Strata/PSUs Unique

When pooling multiple years, each year's strata and PSU IDs must be made globally unique. Otherwise, stratum "1" from 2011 would be treated as the same as stratum "1" from 2019.

```{r pooling-strata, exercise=TRUE}
# Wrong: PSU/strata IDs repeat across years
df_year_check <- df_all |>
  dplyr::group_by(year) |>
  dplyr::summarise(
    n_unique_strata = dplyr::n_distinct(strata),
    n_unique_psu    = dplyr::n_distinct(psu),
    .groups = "drop"
  )
print(df_year_check)
```

```{r make-unique-strata, exercise=TRUE}
# Correct: create year-specific strata and PSU IDs
df_pooled <- df_all |>
  dplyr::mutate(
    strata_id = as.integer(interaction(year, strata, drop = TRUE)),
    psu_id    = as.integer(interaction(year, psu,    drop = TRUE))
  )

# Now strata IDs are unique across years
df_pooled |>
  dplyr::group_by(year) |>
  dplyr::summarise(
    n_unique_strata = dplyr::n_distinct(strata_id),
    .groups = "drop"
  )
```

---

## Exercise: Build the Pooled Design

```{r build-design, exercise=TRUE}
options(survey.lonely.psu = "adjust")

# Build the pooled survey design object
des_all <- survey::svydesign(
  id      = ~psu_id,
  strata  = ~strata_id,
  weights = ~wt,
  data    = df_all,    # already has year-specific strata_id and psu_id
  nest    = TRUE
)

# Inspect the design object
summary(des_all)
```

```{r design-structure, exercise=TRUE}
options(survey.lonely.psu = "adjust")
des_all <- survey::svydesign(
  id = ~psu_id, strata = ~strata_id, weights = ~wt,
  data = df_all, nest = TRUE
)

cat("Class:", class(des_all), "\n")
cat("Number of observations:", nrow(des_all$variables), "\n")
cat("Number of strata:", dplyr::n_distinct(des_all$variables$strata_id), "\n")
cat("Number of PSUs:", dplyr::n_distinct(des_all$variables$psu_id), "\n")
cat("Weight range:", round(range(weights(des_all))), "\n")
```

---

## Updating the Design with New Variables

After building the design, you can add derived variables using `stats::update()`. This is preferable to rebuilding the design from scratch — it keeps the design and data in sync.

```{r update-design, exercise=TRUE}
options(survey.lonely.psu = "adjust")
des_all <- survey::svydesign(
  id = ~psu_id, strata = ~strata_id, weights = ~wt,
  data = df_all, nest = TRUE
)

# Add derived variables to the design
des_all <- stats::update(
  des_all,
  year_lin  = as.numeric(year) - 2011,      # centered time variable
  is_current = dplyr::case_when(
    alcday5 %in% c(777, 999) ~ NA,
    alcday5 == 888 ~ FALSE,
    TRUE ~ TRUE
  ),
  any_binge_f = factor(any_binge_cdc, levels = c(0, 1), labels = c("No", "Yes"))
)

# Verify the new variables are in the design
cat("New variables available:\n")
cat("  year_lin range:", range(des_all$variables$year_lin), "\n")
cat("  is_current distribution:\n")
print(table(des_all$variables$is_current, useNA = "ifany"))
```

---

## Creating Subsets

To analyze current drinkers separately, use `subset()` on the design object — never subset the data frame before building the design.

```{r subset-design, exercise=TRUE}
options(survey.lonely.psu = "adjust")
des_all <- survey::svydesign(
  id = ~psu_id, strata = ~strata_id, weights = ~wt,
  data = df_all, nest = TRUE
)
des_all <- stats::update(des_all,
  is_current = dplyr::case_when(
    alcday5 %in% c(777, 999) ~ NA, alcday5 == 888 ~ FALSE, TRUE ~ TRUE
  )
)

# Subset to current drinkers
des_curr <- subset(des_all, is_current)

cat("All adults design - N:", nrow(des_all$variables), "\n")
cat("Current drinkers design - N:", nrow(des_curr$variables), "\n")
cat("Proportion current drinkers: ",
    round(nrow(des_curr$variables) / nrow(des_all$variables), 3), "\n")
```

```{r subset-quiz}
quiz(
  question(
    "Why do we use subset() on the design object rather than filtering the data frame before building the design?",
    answer("It is faster"),
    answer("Subsetting the design object preserves the correct variance estimation, accounting for the fact that non-selected observations still affect the design structure", correct = TRUE),
    answer("subset() is the only function that works with svydesign objects"),
    answer("It produces the same result either way"),
    allow_retry = TRUE
  )
)
```

---

## The srvyr Package: Tidyverse-Style Surveys

The `srvyr` package wraps `survey` with a dplyr-like interface. This can be cleaner for simple summaries:

```{r srvyr-demo, exercise=TRUE}
library(srvyr)
options(survey.lonely.psu = "adjust")

des_all <- survey::svydesign(
  id = ~psu_id, strata = ~strata_id, weights = ~wt,
  data = df_all, nest = TRUE
)

# Tidyverse-style survey summary
as_survey_design(des_all) |>
  dplyr::filter(!is.na(any_binge_cdc)) |>
  dplyr::group_by(year) |>
  dplyr::summarise(
    prev = survey_mean(any_binge_cdc, vartype = "ci", na.rm = TRUE)
  ) |>
  dplyr::slice_head(n = 6)
```

For this analysis, we use the base `survey` package directly for maximum control over variance estimation and standardization.

---

## Year-Specific Designs for Trend Analysis

For age-standardized trend analysis, we loop over years and build individual year designs. This is necessary because `svystandardize()` only works on a single design at a time.

```{r year-loop-design, exercise=TRUE}
options(survey.lonely.psu = "adjust")

# Pattern for year-specific design (used inside map_dfr loops)
build_year_design <- function(df_pooled, yr) {
  df_yr <- df_pooled |> dplyr::filter(year == yr, !is.na(age7))

  survey::svydesign(
    id      = ~psu_id,
    strata  = ~strata_id,
    weights = ~wt,
    data    = df_yr,
    nest    = TRUE
  )
}

# Build designs for two years
des_2018 <- build_year_design(df_all, 2018)
des_2022 <- build_year_design(df_all, 2022)

cat("2018 design N:", nrow(des_2018$variables), "\n")
cat("2022 design N:", nrow(des_2022$variables), "\n")

# Quick unweighted check on binge prevalence (not for publication!)
cat("\nRaw binge proportions (unweighted check):\n")
cat("  2018:", round(mean(des_2018$variables$any_binge_cdc, na.rm = TRUE), 3), "\n")
cat("  2022:", round(mean(des_2022$variables$any_binge_cdc, na.rm = TRUE), 3), "\n")
```

---

## Module Summary

You have learned:

- Using `mean()` or `lm()` directly on BRFSS data gives **wrong standard errors** — always use `survey::svymean()` and related functions
- `svydesign()` requires three components: **PSUs** (`id`), **strata** (`strata`), and **weights** (`weights`)
- `nest = TRUE` is required because BRFSS PSU IDs repeat across strata
- When pooling years, create **year-specific strata and PSU IDs** using `interaction(year, strata)`
- Use `stats::update()` to add derived variables to an existing design
- Use `subset()` on the design (not the data) to create subgroup designs
- Year-specific design objects are needed for age standardization

**Next:** Module 5 covers computing prevalence estimates and confidence intervals with the survey package.
