---
title: "Module 5: Computing Prevalence Estimates and Confidence Intervals"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: "flatly"
runtime: shiny_prerendered
description: "Use svymean() and svyby() to compute survey-weighted prevalence estimates with confidence intervals."
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(survey)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
tutorial_options(exercise.timelimit = 120)
options(survey.lonely.psu = "adjust")

source("synthetic_brfss.R", local = TRUE)

df_all <- make_synthetic_brfss(n_per_year = 1500, years = 2011:2024) |>
  dplyr::mutate(
    any_binge_cdc = dplyr::case_when(
      rfbing5 == 2 ~ 1L, rfbing5 == 1 ~ 0L, TRUE ~ NA_integer_
    ),
    sex_f = factor(sex, levels = c(2, 1), labels = c("Female", "Male")),
    age_group = dplyr::case_when(
      age5 == 1 ~ "18-24", age5 == 2 ~ "25-29", age5 == 3 ~ "30-34",
      age5 == 4 ~ "35-39", age5 == 5 ~ "40-44", age5 == 6 ~ "45-49",
      age5 == 7 ~ "50-54", age5 == 8 ~ "55-59", age5 == 9 ~ "60-64",
      age5 %in% 10:13 ~ "65+", TRUE ~ NA_character_
    ) |> factor(levels = c("18-24","25-29","30-34","35-39","40-44",
                            "45-49","50-54","55-59","60-64","65+"), ordered = TRUE),
    age7 = dplyr::case_when(
      age5 %in% 1:2 ~ "18-29", age5 %in% 3:4 ~ "30-39",
      age5 %in% 5:6 ~ "40-49", age5 %in% 7:8 ~ "50-59",
      age5 %in% 9:10 ~ "60-69", age5 %in% 11:12 ~ "70-79",
      age5 == 13 ~ "80+", TRUE ~ NA_character_
    ) |> factor(levels = c("18-29","30-39","40-49","50-59","60-69","70-79","80+")),
    educ4 = dplyr::recode(as.integer(educag),
      `1`="<HS", `2`="HS/GED", `3`="Some college/AA", `4`="College",
      .default = NA_character_
    ) |> factor(levels = c("<HS","HS/GED","Some college/AA","College"), ordered = TRUE),
    strata_id = as.integer(interaction(year, strata, drop = TRUE)),
    psu_id    = as.integer(interaction(year, psu,    drop = TRUE)),
    is_current = dplyr::case_when(
      alcday5 %in% c(777, 999) ~ NA, alcday5 == 888 ~ FALSE, TRUE ~ TRUE
    )
  ) |>
  dplyr::filter(!is.na(sex_f), !is.na(age_group))

des_all <- survey::svydesign(
  id = ~psu_id, strata = ~strata_id, weights = ~wt, data = df_all, nest = TRUE
)
des_curr <- subset(des_all, is_current == TRUE)
```

## `svymean()`: Overall Prevalence

`svymean()` computes a weighted mean from a survey design. For a binary outcome (0/1), this gives **prevalence**.

```{r svymean-basic, exercise=TRUE}
# Overall binge drinking prevalence — all years pooled
overall <- survey::svymean(~any_binge_cdc, des_all, na.rm = TRUE)

cat("Prevalence estimate:", round(coef(overall) * 100, 1), "%\n")
cat("Standard error:     ", round(SE(overall) * 100, 2), "pp\n")

# 95% confidence interval
ci <- confint(overall)
cat("95% CI:             ", round(ci[1] * 100, 1), "–", round(ci[2] * 100, 1), "%\n")
```

### Formatting Results

```{r format-helpers, exercise=TRUE}
# Formatting helpers used throughout the analysis
fmt_pct1 <- function(x) sprintf("%.1f", x * 100)
fmt_ci1  <- function(p, l, u) sprintf("%.1f (%.1f-%.1f)", p * 100, l * 100, u * 100)

overall <- survey::svymean(~any_binge_cdc, des_all, na.rm = TRUE)
ci      <- confint(overall)
p       <- coef(overall)[1]
l       <- ci[1, 1]
u       <- ci[1, 2]

cat("Formatted: ", fmt_ci1(p, l, u), "%\n")
```

---

## `svyby()`: Stratified Estimates

`svyby()` computes estimates within subgroups. It's the workhorse for demographic stratification.

```{r svyby-sex, exercise=TRUE}
# Prevalence by sex
prev_sex <- survey::svyby(
  ~any_binge_cdc,    # outcome formula
  ~sex_f,            # subgroup formula
  des_all,           # design object
  survey::svymean,   # statistic function
  na.rm = TRUE,
  vartype = "ci"     # return CI instead of SE
)

prev_sex |>
  tibble::as_tibble() |>
  dplyr::transmute(
    sex       = sex_f,
    prev_pct  = round(any_binge_cdc * 100, 1),
    lcl_pct   = round(ci_l * 100, 1),
    ucl_pct   = round(ci_u * 100, 1),
    ci_fmt    = sprintf("%.1f (%.1f-%.1f)", prev_pct, lcl_pct, ucl_pct)
  )
```

```{r svyby-age, exercise=TRUE}
# Prevalence by age group
prev_age <- survey::svyby(
  ~any_binge_cdc, ~age_group, des_all, survey::svymean,
  na.rm = TRUE, vartype = "ci"
)

prev_age |>
  tibble::as_tibble() |>
  dplyr::transmute(
    age_group = age_group,
    prev_pct  = round(any_binge_cdc * 100, 1),
    lcl_pct   = round(ci_l * 100, 1),
    ucl_pct   = round(ci_u * 100, 1)
  ) |>
  print()
```

---

## Annual Trend: Looping Over Years

For trend analysis, we compute the estimate separately for each year using `purrr::map_dfr()`:

```{r annual-trend, exercise=TRUE}
# Annual prevalence — looped by year
prev_by_year <- purrr::map_dfr(2011:2024, function(yr) {
  # Filter to this year and rebuild design
  df_yr  <- des_all$variables |> dplyr::filter(year == yr)
  des_yr <- survey::svydesign(
    id = ~psu_id, strata = ~strata_id, weights = ~wt,
    data = df_yr, nest = TRUE
  )

  out <- survey::svymean(~any_binge_cdc, des_yr, na.rm = TRUE)
  ci  <- suppressWarnings(confint(out))

  tibble::tibble(
    year = as.integer(yr),
    est  = as.numeric(out)[1],
    lcl  = ci[1, 1],
    ucl  = ci[1, 2]
  )
}) |>
  dplyr::arrange(year)

# Display results
prev_by_year |>
  dplyr::mutate(
    est_pct = round(est * 100, 1),
    ci = sprintf("%.1f (%.1f-%.1f)", est_pct, lcl * 100, ucl * 100)
  ) |>
  dplyr::select(year, ci) |>
  print()
```

---

## Exercise: Stratified Annual Trend

```{r annual-by-sex, exercise=TRUE}
# Compute annual prevalence by sex
prev_sex_byyear <- purrr::map_dfr(2011:2024, function(yr) {
  df_yr  <- des_all$variables |> dplyr::filter(year == yr)
  des_yr <- survey::svydesign(
    id = ~psu_id, strata = ~strata_id, weights = ~wt,
    data = df_yr, nest = TRUE
  )

  survey::svyby(
    ~any_binge_cdc, ~sex_f, des_yr, survey::svymean,
    na.rm = TRUE, vartype = "ci"
  ) |>
    tibble::as_tibble() |>
    dplyr::mutate(year = yr)
}) |>
  dplyr::transmute(
    sex  = as.character(sex_f),
    year = as.integer(year),
    est  = any_binge_cdc,
    lcl  = ci_l,
    ucl  = ci_u
  ) |>
  dplyr::arrange(sex, year)

# Show first 10 rows
head(prev_sex_byyear, 10)
```

---

## Adjacent-Year Differences

Adjacent-year differences test whether prevalence changed significantly from one year to the next. We compute the difference and its standard error using the delta method:

```{r adjacent-diffs}
# Compute adjacent-year differences with SE
compute_adjacent_diffs <- function(df_prev) {
  df_prev |>
    dplyr::arrange(year) |>
    dplyr::mutate(
      prev_est = dplyr::lag(est),
      prev_lcl = dplyr::lag(lcl),
      prev_ucl = dplyr::lag(ucl),
      delta_pp = (est - prev_est) * 100,
      # Approximate SE from CI width (conservative)
      se_curr  = (ucl - lcl) / (2 * 1.96),
      se_prev  = (dplyr::lag(ucl) - dplyr::lag(lcl)) / (2 * 1.96),
      se_delta = sqrt(se_curr^2 + se_prev^2),
      z        = (est - prev_est) / se_delta,
      p_value  = 2 * (1 - pnorm(abs(z)))
    ) |>
    dplyr::filter(!is.na(prev_est))
}
```

```{r adjacent-diffs-exercise, exercise=TRUE}
prev_by_year <- purrr::map_dfr(2011:2024, function(yr) {
  df_yr  <- des_all$variables |> dplyr::filter(year == yr)
  des_yr <- survey::svydesign(
    id = ~psu_id, strata = ~strata_id, weights = ~wt,
    data = df_yr, nest = TRUE
  )
  out <- survey::svymean(~any_binge_cdc, des_yr, na.rm = TRUE)
  ci  <- suppressWarnings(confint(out))
  tibble::tibble(year = as.integer(yr), est = as.numeric(out)[1],
                 lcl = ci[1,1], ucl = ci[1,2])
}) |> dplyr::arrange(year)

diffs <- compute_adjacent_diffs(prev_by_year)

diffs |>
  dplyr::transmute(
    years     = paste0(year - 1, "-", year),
    delta_pp  = round(delta_pp, 2),
    p_value   = round(p_value, 3),
    sig       = dplyr::case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      TRUE            ~ ""
    )
  ) |>
  print()
```

---

## Multiple Comparisons: Benjamini-Hochberg Correction

When testing many adjacent-year differences simultaneously, we apply the **Benjamini-Hochberg (BH) false discovery rate** correction to control for multiple comparisons.

```{r bh-correction, exercise=TRUE}
prev_by_year <- purrr::map_dfr(2011:2024, function(yr) {
  df_yr  <- des_all$variables |> dplyr::filter(year == yr)
  des_yr <- survey::svydesign(
    id = ~psu_id, strata = ~strata_id, weights = ~wt,
    data = df_yr, nest = TRUE
  )
  out <- survey::svymean(~any_binge_cdc, des_yr, na.rm = TRUE)
  ci  <- suppressWarnings(confint(out))
  tibble::tibble(year = as.integer(yr), est = as.numeric(out)[1],
                 lcl = ci[1,1], ucl = ci[1,2])
}) |> dplyr::arrange(year)

diffs <- compute_adjacent_diffs(prev_by_year) |>
  dplyr::mutate(
    p_adj_bh = p.adjust(p_value, method = "BH"),
    sig_bh   = dplyr::case_when(
      p_adj_bh < 0.05 ~ "*",
      TRUE            ~ ""
    )
  )

diffs |>
  dplyr::transmute(
    years     = paste0(year - 1, "-", year),
    delta_pp  = round(delta_pp, 2),
    p_raw     = round(p_value, 3),
    p_adj_bh  = round(p_adj_bh, 3),
    sig_bh
  ) |>
  print()
```

```{r bh-quiz}
quiz(
  question(
    "Why use Benjamini-Hochberg correction rather than Bonferroni when testing 13 adjacent-year differences?",
    answer("BH is always more powerful than Bonferroni"),
    answer("Bonferroni controls the family-wise error rate (very conservative); BH controls the false discovery rate, which is more appropriate when many tests are expected to be true positives", correct = TRUE),
    answer("BH is the only method available in base R"),
    answer("The tests are independent, so BH is required"),
    allow_retry = TRUE
  )
)
```

---

## Exporting Results

Results are exported in two formats: proportions (0–1) and percentages (0–100).

```{r export-pattern, eval=FALSE}
# Export proportions (for programmatic use)
readr::write_csv(
  prev_by_year,
  file.path(tab_dir, "main_prev_any_binge_alladults_CI_prop.csv")
)

# Export percentages (human-readable)
readr::write_csv(
  prev_by_year |>
    dplyr::mutate(
      est = est * 100,
      lcl = lcl * 100,
      ucl = ucl * 100,
      ci  = sprintf("%.1f (%.1f-%.1f)", est, lcl, ucl)
    ),
  file.path(tab_dir, "main_prev_any_binge_alladults_CI_pct.csv")
)
```

The `ci` column bundles the estimate and CI into a single publication-ready string like `"18.2 (17.9-18.5)"`.

---

## Module Summary

You have learned:

- `svymean()` computes a weighted prevalence with correct survey-design standard errors
- `svyby()` computes estimates within subgroups (sex, age, race, education)
- For trend analysis, **loop over years** building a fresh design for each year
- `p.adjust(method = "BH")` applies Benjamini-Hochberg correction for multiple adjacent-year comparisons
- Export results as both proportions (for computation) and percentages (for publication)

**Next:** Module 6 covers age standardization using the Year 2000 standard population — essential for valid trend comparisons when population age structure changes over time.
