---
title: "Module 6: Age Standardization"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: "flatly"
runtime: shiny_prerendered
description: "Apply direct age standardization using the Year 2000 standard population and svystandardize()."
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(survey)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
tutorial_options(exercise.timelimit = 120)
options(survey.lonely.psu = "adjust")

source("synthetic_brfss.R", local = TRUE)

df_all <- make_synthetic_brfss(n_per_year = 1500, years = 2011:2024) |>
  dplyr::mutate(
    any_binge_cdc = dplyr::case_when(
      rfbing5 == 2 ~ 1L, rfbing5 == 1 ~ 0L, TRUE ~ NA_integer_
    ),
    sex_f = factor(sex, levels = c(2, 1), labels = c("Female", "Male")),
    age7 = dplyr::case_when(
      age5 %in% 1:2 ~ "18-29", age5 %in% 3:4 ~ "30-39",
      age5 %in% 5:6 ~ "40-49", age5 %in% 7:8 ~ "50-59",
      age5 %in% 9:10 ~ "60-69", age5 %in% 11:12 ~ "70-79",
      age5 == 13 ~ "80+", TRUE ~ NA_character_
    ) |> factor(levels = c("18-29","30-39","40-49","50-59","60-69","70-79","80+")),
    strata_id = as.integer(interaction(year, strata, drop = TRUE)),
    psu_id    = as.integer(interaction(year, psu,    drop = TRUE))
  ) |>
  dplyr::filter(!is.na(sex_f), !is.na(age7))

# Year 2000 standard population weights (7-band, adults 18+)
std2000_5yr <- c(
  "15-19"=72169, "20-24"=66478, "25-29"=64529, "30-34"=71044, "35-39"=80762,
  "40-44"=81851, "45-49"=72118, "50-54"=62716, "55-59"=48454, "60-64"=38793,
  "65-69"=34264, "70-74"=31773, "75-79"=26999, "80-84"=17842, "85+"=15508
)

g <- function(k) std2000_5yr[[k]]

std_10bin_counts <- c(
  "18-24" = 0.4*g("15-19") + g("20-24"),
  "25-29" = g("25-29"), "30-34" = g("30-34"), "35-39" = g("35-39"),
  "40-44" = g("40-44"), "45-49" = g("45-49"), "50-54" = g("50-54"),
  "55-59" = g("55-59"), "60-64" = g("60-64"),
  "65+" = g("65-69") + g("70-74") + g("75-79") + g("80-84") + g("85+")
)

w_2000_age7_counts <- c(
  "18-29" = unname(std_10bin_counts["18-24"]) + unname(std_10bin_counts["25-29"]),
  "30-39" = unname(std_10bin_counts["30-34"]) + unname(std_10bin_counts["35-39"]),
  "40-49" = unname(std_10bin_counts["40-44"]) + unname(std_10bin_counts["45-49"]),
  "50-59" = unname(std_10bin_counts["50-54"]) + unname(std_10bin_counts["55-59"]),
  "60-69" = unname(std_10bin_counts["60-64"]) + g("65-69"),
  "70-79" = g("70-74") + g("75-79"),
  "80+"   = g("80-84") + g("85+")
)

w_2000_age7 <- w_2000_age7_counts / sum(w_2000_age7_counts)
```

## Why Age-Standardize?

Binge drinking rates differ sharply by age. If the **age distribution of the U.S. population changes over time** (e.g., the population is aging), then crude prevalence estimates will trend downward simply because older people drink less — not because actual behavior changed.

**Age standardization removes this confounding** by applying a fixed reference age distribution to all years. This lets us compare "what would prevalence be if all years had the same age structure?"

**Example:**

```{r why-standardize, exercise=TRUE}
# Demonstrate the effect of age structure
# Suppose younger age groups have higher binge rates
# If 2011 had more young adults than 2024, crude rates would appear higher in 2011
# even if age-specific rates were identical

age_rates <- tibble::tibble(
  age_group = c("18-29", "30-39", "40-49", "50-59", "60-69", "70-79", "80+"),
  binge_rate = c(0.35,    0.25,    0.18,    0.12,    0.07,    0.04,    0.02)
)

# Younger population structure (2011-like)
pop_young <- c(0.22, 0.20, 0.18, 0.16, 0.13, 0.07, 0.04)
# Older population structure (2024-like)
pop_older <- c(0.17, 0.17, 0.18, 0.17, 0.16, 0.09, 0.06)

crude_young <- sum(age_rates$binge_rate * pop_young)
crude_older <- sum(age_rates$binge_rate * pop_older)

cat("Crude rate (young pop structure):", round(crude_young * 100, 1), "%\n")
cat("Crude rate (older pop structure):", round(crude_older * 100, 1), "%\n")
cat("Difference:", round((crude_young - crude_older) * 100, 1), "pp\n\n")
cat("This difference is ENTIRELY due to age structure change —\n")
cat("not any actual change in drinking behavior!\n")
```

---

## The Year 2000 Standard Population

The U.S. uses the **Year 2000 U.S. standard population** (from NCHS Statistical Note 20) as the reference for age standardization. This is the official CDC standard and allows comparison with other studies and surveillance data.

```{r year2000-std}
# Year 2000 standard population counts (5-year bands, in thousands)
std2000_5yr <- c(
  "15-19" = 72169, "20-24" = 66478, "25-29" = 64529,
  "30-34" = 71044, "35-39" = 80762, "40-44" = 81851,
  "45-49" = 72118, "50-54" = 62716, "55-59" = 48454,
  "60-64" = 38793, "65-69" = 34264, "70-74" = 31773,
  "75-79" = 26999, "80-84" = 17842, "85+"   = 15508
)
```

We collapse these 5-year bands into **7 age bands** that match the BRFSS `_AGEG5YR` structure:

```{r age7-weights, exercise=TRUE}
g <- function(k) std2000_5yr[[k]]

std_10bin_counts <- c(
  "18-24" = 0.4 * g("15-19") + g("20-24"),  # 18-19 = 2/5 of 15-19 band
  "25-29" = g("25-29"), "30-34" = g("30-34"), "35-39" = g("35-39"),
  "40-44" = g("40-44"), "45-49" = g("45-49"), "50-54" = g("50-54"),
  "55-59" = g("55-59"), "60-64" = g("60-64"),
  "65+"   = g("65-69") + g("70-74") + g("75-79") + g("80-84") + g("85+")
)

# Combine into 7 bands
w_2000_age7_counts <- c(
  "18-29" = unname(std_10bin_counts["18-24"]) + unname(std_10bin_counts["25-29"]),
  "30-39" = unname(std_10bin_counts["30-34"]) + unname(std_10bin_counts["35-39"]),
  "40-49" = unname(std_10bin_counts["40-44"]) + unname(std_10bin_counts["45-49"]),
  "50-59" = unname(std_10bin_counts["50-54"]) + unname(std_10bin_counts["55-59"]),
  "60-69" = unname(std_10bin_counts["60-64"]) + g("65-69"),
  "70-79" = g("70-74") + g("75-79"),
  "80+"   = g("80-84") + g("85+")
)

# Proportional weights (must sum to 1)
w_2000_age7 <- w_2000_age7_counts / sum(w_2000_age7_counts)
round(w_2000_age7 * 100, 1)  # as percentages
```

Note how the 18–29 and 30–39 bands receive the most weight — reflecting the relatively young Year 2000 U.S. population.

```{r weights-quiz}
quiz(
  question(
    "Why do we use counts (not proportions) in svystandardize()?",
    answer("Counts produce more accurate estimates"),
    answer("svystandardize() internally converts counts to proportions, so either works"),
    answer("The survey package requires population counts, not proportions, to scale the standardization weights correctly", correct = TRUE),
    answer("Using proportions would change the point estimate"),
    allow_retry = TRUE
  )
)
```

---

## `svystandardize()`: Direct Age Standardization

The `survey::svystandardize()` function modifies a survey design so that the resulting estimates are age-standardized to a reference population.

```{r svystandardize-demo, exercise=TRUE}
# Age-standardized prevalence for year 2019
df_yr <- df_all |> dplyr::filter(year == 2019, !is.na(age7))

des_yr <- survey::svydesign(
  id = ~psu_id, strata = ~strata_id, weights = ~wt,
  data = df_yr, nest = TRUE
)

# Apply direct standardization to Year 2000 7-band standard
des_std <- survey::svystandardize(
  des_yr,
  by         = ~age7,               # standardize over this variable
  population = w_2000_age7_counts   # reference population counts
)

# Compute age-adjusted prevalence
out_adj <- survey::svymean(~any_binge_cdc, des_std, na.rm = TRUE)
out_crude <- survey::svymean(~any_binge_cdc, des_yr, na.rm = TRUE)

ci_adj   <- suppressWarnings(confint(out_adj))
ci_crude <- suppressWarnings(confint(out_crude))

cat("Year 2019:\n")
cat(sprintf("  Crude:        %.1f%% (%.1f-%.1f)\n",
    coef(out_crude)[1]*100, ci_crude[1]*100, ci_crude[2]*100))
cat(sprintf("  Age-adjusted: %.1f%% (%.1f-%.1f)\n",
    coef(out_adj)[1]*100, ci_adj[1]*100, ci_adj[2]*100))
```

---

## Age-Standardized Trend: Full 2011–2024 Loop

```{r ageadj-trend, exercise=TRUE}
prev_all_prop_ageadj <- purrr::map_dfr(2011:2024, function(yr) {
  df_yr <- df_all |> dplyr::filter(year == yr, !is.na(age7))

  des_yr <- survey::svydesign(
    id = ~psu_id, strata = ~strata_id, weights = ~wt,
    data = df_yr, nest = TRUE
  )

  # Apply Year 2000 age standardization
  des_std <- survey::svystandardize(
    des_yr,
    by         = ~age7,
    population = w_2000_age7_counts
  )

  out <- survey::svymean(~any_binge_cdc, des_std, na.rm = TRUE)
  ci  <- suppressWarnings(confint(out))

  tibble::tibble(
    year = as.integer(yr),
    est  = as.numeric(out)[1],
    lcl  = ci[1, 1],
    ucl  = ci[1, 2]
  )
}) |>
  dplyr::arrange(year)

prev_all_prop_ageadj |>
  dplyr::mutate(
    est_pct = round(est * 100, 1),
    ci = sprintf("%.1f (%.1f-%.1f)", est_pct, lcl * 100, ucl * 100)
  ) |>
  dplyr::select(year, ci) |>
  print()
```

---

## Stratified Age-Adjusted Estimates

To get age-adjusted estimates *within* demographic groups (e.g., by sex), use the `over` argument in `svystandardize()`:

```{r strat-ageadj, exercise=TRUE}
# Age-adjusted prevalence by sex — 2022
df_yr <- df_all |> dplyr::filter(year == 2022, !is.na(age7), !is.na(sex_f))

des_yr <- survey::svydesign(
  id = ~psu_id, strata = ~strata_id, weights = ~wt,
  data = df_yr, nest = TRUE
)

# Standardize within each sex group separately
des_std <- survey::svystandardize(
  des_yr,
  by         = ~age7,    # age bands to standardize over
  over       = ~sex_f,   # compute separately for each sex
  population = w_2000_age7_counts
)

out <- survey::svyby(
  ~any_binge_cdc, ~sex_f, des_std, survey::svymean,
  na.rm = TRUE, vartype = "ci"
)

out |>
  tibble::as_tibble() |>
  dplyr::transmute(
    sex      = sex_f,
    prev_pct = round(any_binge_cdc * 100, 1),
    lcl_pct  = round(ci_l * 100, 1),
    ucl_pct  = round(ci_u * 100, 1),
    ci       = sprintf("%.1f (%.1f-%.1f)", prev_pct, lcl_pct, ucl_pct)
  )
```

---

## Crude vs. Age-Adjusted: Side-by-Side Comparison

```{r crude-vs-adj, exercise=TRUE}
comparison <- purrr::map_dfr(c(2011, 2015, 2019, 2024), function(yr) {
  df_yr <- df_all |> dplyr::filter(year == yr, !is.na(age7))
  des_yr <- survey::svydesign(
    id = ~psu_id, strata = ~strata_id, weights = ~wt,
    data = df_yr, nest = TRUE
  )
  des_std <- survey::svystandardize(des_yr, by = ~age7, population = w_2000_age7_counts)

  crude <- coef(survey::svymean(~any_binge_cdc, des_yr, na.rm = TRUE))[1]
  adj   <- coef(survey::svymean(~any_binge_cdc, des_std, na.rm = TRUE))[1]

  tibble::tibble(year = yr, crude_pct = round(crude * 100, 1), adj_pct = round(adj * 100, 1))
})

comparison |>
  dplyr::mutate(diff_pp = round(adj_pct - crude_pct, 1)) |>
  print()
```

---

## Module Summary

You have learned:

- **Why age-standardize:** Crude rates are confounded by population age structure; standardization enables valid time-trend comparisons
- **Year 2000 standard:** The U.S. CDC standard; creates 7 bands from NCHS 5-year population counts
- **`svystandardize()`:** Directly standardizes a survey design to a reference population; returns a modified design object that produces age-adjusted estimates
- **`over` argument:** Compute stratified age-adjusted estimates simultaneously for demographic groups
- Age-adjusted estimates will differ from crude estimates when the study population's age distribution differs from the standard

**Next:** Module 7 covers stratified trend analysis across demographic subgroups and formal trend testing.
