---
title: "Module 8: Visualization"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: "flatly"
runtime: shiny_prerendered
description: "Build publication-quality trend plots with confidence interval ribbons, annotations, and facets using ggplot2."
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(survey)
library(scales)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 8, fig.height = 4.5)
tutorial_options(exercise.timelimit = 120)
options(survey.lonely.psu = "adjust")

source("synthetic_brfss.R", local = TRUE)

# Build pre-computed prevalence data for plotting
df_all <- make_synthetic_brfss(n_per_year = 1000, years = 2011:2024) |>
  dplyr::mutate(
    any_binge_cdc = dplyr::case_when(rfbing5 == 2 ~ 1L, rfbing5 == 1 ~ 0L, TRUE ~ NA_integer_),
    sex_f = factor(sex, levels = c(2, 1), labels = c("Female", "Male")),
    age7 = dplyr::case_when(
      age5 %in% 1:2 ~ "18-29", age5 %in% 3:4 ~ "30-39", age5 %in% 5:6 ~ "40-49",
      age5 %in% 7:8 ~ "50-59", age5 %in% 9:10 ~ "60-69", age5 %in% 11:12 ~ "70-79",
      age5 == 13 ~ "80+", TRUE ~ NA_character_
    ) |> factor(levels = c("18-29","30-39","40-49","50-59","60-69","70-79","80+")),
    strata_id = as.integer(interaction(year, strata, drop = TRUE)),
    psu_id    = as.integer(interaction(year, psu,    drop = TRUE))
  ) |> dplyr::filter(!is.na(sex_f), !is.na(age7))

g <- function(k) c("15-19"=72169,"20-24"=66478,"25-29"=64529,"30-34"=71044,"35-39"=80762,
                   "40-44"=81851,"45-49"=72118,"50-54"=62716,"55-59"=48454,"60-64"=38793,
                   "65-69"=34264,"70-74"=31773,"75-79"=26999,"80-84"=17842,"85+"=15508)[[k]]

w_2000_age7_counts <- c(
  "18-29" = 0.4*g("15-19") + g("20-24") + g("25-29"),
  "30-39" = g("30-34") + g("35-39"), "40-49" = g("40-44") + g("45-49"),
  "50-59" = g("50-54") + g("55-59"), "60-69" = g("60-64") + g("65-69"),
  "70-79" = g("70-74") + g("75-79"), "80+" = g("80-84") + g("85+")
)

# Compute age-adjusted overall trend
prev_all <- purrr::map_dfr(2011:2024, function(yr) {
  df_yr <- df_all |> dplyr::filter(year == yr)
  des_yr <- survey::svydesign(id=~psu_id, strata=~strata_id, weights=~wt, data=df_yr, nest=TRUE)
  des_std <- survey::svystandardize(des_yr, by=~age7, population=w_2000_age7_counts)
  out <- survey::svymean(~any_binge_cdc, des_std, na.rm=TRUE)
  ci <- suppressWarnings(confint(out))
  tibble::tibble(year=as.integer(yr), est=as.numeric(out)[1], lcl=ci[1,1], ucl=ci[1,2])
}) |> dplyr::arrange(year) |>
  dplyr::mutate(est_pct=est*100, lcl_pct=lcl*100, ucl_pct=ucl*100,
                roll3=(dplyr::lag(est_pct)+est_pct+dplyr::lead(est_pct))/3,
                delta_pp=est_pct - dplyr::lag(est_pct))

# Compute age-adjusted by sex
prev_sex <- purrr::map_dfr(2011:2024, function(yr) {
  df_yr <- df_all |> dplyr::filter(year == yr, !is.na(sex_f))
  des_yr <- survey::svydesign(id=~psu_id, strata=~strata_id, weights=~wt, data=df_yr, nest=TRUE)
  des_std <- survey::svystandardize(des_yr, by=~age7, over=~sex_f, population=w_2000_age7_counts)
  survey::svyby(~any_binge_cdc, ~sex_f, des_std, survey::svymean, na.rm=TRUE, vartype="ci") |>
    tibble::as_tibble() |> dplyr::mutate(year=yr)
}) |>
  dplyr::transmute(sex=as.character(sex_f), year=as.integer(year), est=any_binge_cdc, lcl=ci_l, ucl=ci_u) |>
  dplyr::mutate(est_pct=est*100, lcl_pct=lcl*100, ucl_pct=ucl*100) |>
  dplyr::arrange(sex, year)

# Age-specific prevalence (pooled 2011-2024)
df_all_des <- survey::svydesign(id=~psu_id, strata=~strata_id, weights=~wt, data=df_all, nest=TRUE)
prev_age7 <- survey::svyby(~any_binge_cdc, ~age7, df_all_des, survey::svymean, na.rm=TRUE, vartype="ci") |>
  tibble::as_tibble() |>
  dplyr::transmute(age7=age7, est_pct=any_binge_cdc*100, lcl_pct=ci_l*100, ucl_pct=ci_u*100)
```

## Custom Theme: `theme_klean()`

A minimal, publication-ready ggplot2 theme is used throughout the analysis:

```{r theme-klean-def}
theme_klean <- function(base_size = 12) {
  ggplot2::theme_minimal(base_size = base_size) +
    ggplot2::theme(
      panel.grid.minor    = ggplot2::element_blank(),
      panel.grid.major.x  = ggplot2::element_blank(),
      axis.title.x        = ggplot2::element_blank(),
      plot.title          = ggplot2::element_text(face = "bold"),
      plot.subtitle       = ggplot2::element_text(color = "grey30"),
      plot.caption        = ggplot2::element_text(size = ggplot2::rel(0.9), color = "grey40"),
      plot.margin         = ggplot2::margin(10, 18, 10, 10),
      legend.position     = "none"
    )
}
```

**Design decisions:**
- No minor gridlines (reduce visual clutter)
- No major vertical gridlines (trend lines already convey the x-axis structure)
- No x-axis title (year is self-evident)
- Bold title, muted subtitle and caption

---

## Figure 1: Line + CI Ribbon

The primary trend figure: a solid line for estimates, a ribbon for 95% CI, and a dotted rolling average.

```{r fig1-main, exercise=TRUE}
yr_first <- min(prev_all$year)
yr_last  <- max(prev_all$year)
pp_change <- with(prev_all, est_pct[year == yr_last] - est_pct[year == yr_first])

last_row <- prev_all[prev_all$year == yr_last, , drop = FALSE]
end_lab  <- sprintf("%d: %.1f%% (%.1f-%.1f)", yr_last,
                    last_row$est_pct, last_row$lcl_pct, last_row$ucl_pct)
delta_lab <- sprintf("Change since %d: %+.1f pp", yr_first, pp_change)

p_main <- ggplot(prev_all, aes(x = year, y = est_pct)) +
  # COVID period shading
  annotate("rect",
           xmin = 2019.5, xmax = 2020.5,
           ymin = -Inf,   ymax = Inf,
           fill = "grey90", alpha = 0.4) +
  # CI ribbon
  geom_ribbon(aes(ymin = lcl_pct, ymax = ucl_pct), alpha = 0.18) +
  # Trend line
  geom_line(linewidth = 1.1) +
  # Data points
  geom_point(size = 2.2) +
  # 3-point rolling average
  geom_line(aes(y = roll3), linewidth = 0.8, linetype = "dotted") +
  # End-label annotation
  geom_text(data = last_row, aes(label = end_lab),
            hjust = -0.05, vjust = 0.5, size = 3.8) +
  # Delta annotation
  annotate("text",
           x = yr_last, y = max(prev_all$ucl_pct, na.rm = TRUE),
           label = delta_lab,
           vjust = -0.8, hjust = 1, size = 3.5, color = "grey30") +
  scale_x_continuous(
    breaks = scales::pretty_breaks(),
    expand = expansion(mult = c(0.01, 0.12))
  ) +
  scale_y_continuous(
    labels = scales::label_percent(accuracy = 1, scale = 1),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title    = "Binge drinking prevalence (Any binge, All adults)",
    subtitle = "Age-adjusted to Year-2000 standard (7 bands); BRFSS",
    y        = "Prevalence (%)",
    caption  = "Shaded band = 95% CI. Dotted line = 3-pt rolling average.\nGrey rectangle = COVID period (2020). Synthetic data for illustration."
  ) +
  coord_cartesian(clip = "off") +
  theme_klean()

p_main
```

---

## Figure 2: Dot + CI Error Bar

Cleaner for showing individual-year uncertainty without the visual weight of a ribbon:

```{r fig2-dotci, exercise=TRUE}
p_dotci <- ggplot(prev_all, aes(x = year, y = est_pct)) +
  geom_linerange(aes(ymin = lcl_pct, ymax = ucl_pct),
                 linewidth = 0.9, alpha = 0.7) +
  geom_point(size = 2.2) +
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  scale_y_continuous(
    labels = scales::label_percent(accuracy = 1, scale = 1),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title    = "Binge drinking prevalence (Any binge, All adults)",
    subtitle = "Age-adjusted; dot = point estimate, line = 95% CI",
    y        = "Prevalence (%)"
  ) +
  theme_klean()

p_dotci
```

---

## Figure 3: Year-Over-Year Change (Bar Chart)

Shows the magnitude and direction of each annual change:

```{r fig3-delta, exercise=TRUE}
p_delta <- prev_all |>
  dplyr::filter(!is.na(delta_pp)) |>
  ggplot(aes(x = year, y = delta_pp, fill = delta_pp > 0)) +
  geom_col(width = 0.7, alpha = 0.9, show.legend = FALSE) +
  geom_hline(yintercept = 0, linewidth = 0.6, color = "grey50") +
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  scale_y_continuous(
    labels = function(x) sprintf("%+.1f", x),
    expand = expansion(mult = c(0.05, 0.05))
  ) +
  scale_fill_manual(values = c("TRUE" = "#e15759", "FALSE" = "#4e79a7")) +
  labs(
    title    = "Year-over-year change in binge drinking prevalence",
    subtitle = sprintf("Net change %d-%d: %+.1f pp",
                       min(prev_all$year), max(prev_all$year),
                       with(prev_all,
                            est_pct[year==max(year)] - est_pct[year==min(year)])),
    y        = "Change (percentage points)"
  ) +
  theme_klean()

p_delta
```

---

## Figure 4: Stratified Trend by Sex

```{r fig4-sex, exercise=TRUE}
p_sex <- ggplot(prev_sex, aes(x = year, y = est_pct, color = sex, fill = sex, group = sex)) +
  # COVID period
  annotate("rect", xmin = 2019.5, xmax = 2020.5, ymin = -Inf, ymax = Inf,
           fill = "grey90", alpha = 0.4) +
  geom_ribbon(aes(ymin = lcl_pct, ymax = ucl_pct), alpha = 0.15, color = NA) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  scale_y_continuous(
    labels = scales::label_percent(accuracy = 1, scale = 1),
    expand = expansion(mult = c(0, 0.05))
  ) +
  scale_color_manual(values = c("Female" = "#e15759", "Male" = "#4e79a7")) +
  scale_fill_manual(values  = c("Female" = "#e15759", "Male" = "#4e79a7")) +
  labs(
    title    = "Binge drinking prevalence by sex",
    subtitle = "Age-adjusted to Year-2000 standard; BRFSS",
    y        = "Prevalence (%)",
    color    = NULL, fill = NULL
  ) +
  theme_klean() +
  theme(legend.position = "bottom")

p_sex
```

---

## Figure 5: Age Band — Dot + CI (Pooled)

```{r fig5-age, exercise=TRUE}
p_age7 <- ggplot(prev_age7, aes(x = age7, y = est_pct)) +
  geom_linerange(aes(ymin = lcl_pct, ymax = ucl_pct), linewidth = 1) +
  geom_point(size = 2.4) +
  scale_y_continuous(
    labels = scales::label_percent(accuracy = 1, scale = 1),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title    = "Any binge drinking by age band (pooled 2011-2024)",
    subtitle = "BRFSS; survey-weighted; 95% CI",
    y        = "Prevalence (%)",
    x        = "Age band",
    caption  = "Note: not age-standardized — age is the stratifier."
  ) +
  theme_klean() +
  theme(axis.title.x = element_text())

p_age7
```

---

## Figure 6: Faceted Small Multiples (Age × Year)

Faceted plots are useful when you have many subgroups — one panel per subgroup keeps trends readable:

```{r fig6-facet, exercise=TRUE}
# Age-specific trends by year
prev_age_byyear <- purrr::map_dfr(2011:2024, function(yr) {
  df_yr <- df_all |> dplyr::filter(year == yr)
  des_yr <- survey::svydesign(id=~psu_id, strata=~strata_id, weights=~wt, data=df_yr, nest=TRUE)
  survey::svyby(~any_binge_cdc, ~age7, des_yr, survey::svymean, na.rm=TRUE, vartype="ci") |>
    tibble::as_tibble() |> dplyr::mutate(year = yr)
}) |>
  dplyr::transmute(
    age7 = factor(age7, levels = c("18-29","30-39","40-49","50-59","60-69","70-79","80+")),
    year = as.integer(year),
    est_pct = any_binge_cdc * 100,
    lcl_pct = ci_l * 100,
    ucl_pct = ci_u * 100
  )

p_facet <- ggplot(prev_age_byyear, aes(x = year, y = est_pct)) +
  geom_ribbon(aes(ymin = lcl_pct, ymax = ucl_pct), alpha = 0.18) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.9) +
  facet_wrap(~ age7, ncol = 4) +
  scale_x_continuous(breaks = c(2011, 2017, 2024)) +
  scale_y_continuous(
    labels = scales::label_percent(accuracy = 1, scale = 1)
  ) +
  labs(
    title    = "Any binge drinking by age band over time",
    subtitle = "BRFSS; survey-weighted; 95% CI ribbons",
    y        = "Prevalence (%)"
  ) +
  theme_klean() +
  theme(axis.title.x = ggplot2::element_text())

p_facet
```

---

## Figure 7: Heatmap (Year × Age)

Heatmaps reveal patterns quickly across a two-dimensional table:

```{r fig7-heatmap, exercise=TRUE}
p_heat <- ggplot(prev_age_byyear, aes(x = year, y = age7, fill = est_pct)) +
  geom_tile() +
  scale_fill_gradient(
    name = "Prev (%)",
    low  = "white",
    high = "black"
  ) +
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  labs(
    title    = "Any binge drinking: heatmap by year and age band",
    subtitle = "Darker = higher prevalence; survey-weighted",
    x        = NULL, y = NULL
  ) +
  theme_klean() +
  theme(legend.position = "right")

p_heat
```

---

## Saving Figures

The `save_both()` helper saves every figure as both PNG (for reports/presentations) and SVG (for vector editing):

```{r save-both, eval=FALSE}
save_both <- function(p, fname, w = 8, h = 4.75, dpi = 320) {
  ggplot2::ggsave(
    file.path(fig_dir, paste0(fname, ".png")),
    p, width = w, height = h, units = "in", dpi = dpi, bg = "white"
  )
  ggplot2::ggsave(
    file.path(fig_dir, paste0(fname, ".svg")),
    p, width = w, height = h, units = "in", dpi = dpi, bg = "white"
  )
}

# Example usage:
save_both(p_main, "main_prev_any_binge_alladults_AGEADJ2000_line_ribbon")
save_both(p_sex,  "main_prev_any_binge_by_sex_AGEADJ2000_line_ribbon")
```

**Key settings:**
- `dpi = 320` — exceeds journal minimum (300 dpi) with room for layout scaling
- `bg = "white"` — prevents transparent backgrounds in PNG
- Fixed dimensions in inches (`units = "in"`) — consistent sizing across figures

---

## Module Summary

You have learned:

- `theme_klean()` creates a minimal, publication-quality ggplot2 theme
- `geom_ribbon()` shows 95% CI as a shaded band around the trend line
- `annotate("rect", ...)` highlights specific periods (e.g., COVID, policy changes)
- Rolling averages (`lag()` + `lead()`) smooth noisy estimates for visual clarity
- `facet_wrap()` creates small-multiple panels for multi-subgroup trends
- `geom_tile()` + `scale_fill_gradient()` builds heatmaps for two-dimensional patterns
- `save_both()` exports PNG (print) and SVG (vector) simultaneously

**Next:** Module 9 covers regression modeling — Prevalence Ratio (PR) models, interaction terms, and effect modification testing.
